Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-28 draft.tex
completed Steffens correction

\documentclass[journal]{style/vgtc} 			          % final (journal style)
%\documentclass[review,journal]{style/vgtc}         % review (journal style)
%\documentclass[widereview]{style/vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{style/vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{associations}).
%
Modern cohort studies also incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It shows to be a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.

@@

% \abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
% %
% Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
% %
% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{associations}).
% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{associations}).
% %
% Modern cohort studies also incorporate medical image data.
% %
% Analyzing these data requires image segmentation, extraction of key figures, and shape based subject grouping.
% %
% The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
% \\\\
% We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
% %
% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.
% %As extension to the epidemiological workflow it aims to unveil associations based on all reliable shape information rather than abstracting it to a few key figures.
% %Iterative analysis between ... allows for extracting associations which without the visiual analysis driven approach maybe would be overlooked. %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen 
% %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
% %Iterative analysis of image-based epidemiological datasets aims to provide insight into associations between organ shape and % in order to provide insights into \emph{associations} by incorporating additional information
% %It extends the epidemiological workflow with an iterative Analysis process and 
% It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
% %
% Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
% %
% Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
% %
% Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
% } % end of abstract

@@

\paragraph{Domain and Range Variables. }
In the \emph{IVA} context, data is divided into two major view types.
%
The human body exposing shape information for the \emph{physical view} \cite{Oeltze2013}.
%
This information space is usually displayed via volume rendering techniques \cite{Oeltze2007}.
%
%Spatiotemporal views include data from the \emph{Object Space Domain} renders data centered to the focused domain, in our example the medical image data space.
%
These variables spanning the 3D-space are also referred to as \emph{domain variables}.
%
\emph{Range variables} in the epidemiological context can be divided twofold:
%
\begin{itemize}
	\item Variables derived from the image data. 
	%
	These measures quantify shape information to allow for comparison and can also be used to brush in the image space.
	%
	%These variables describe image data and can also be used to brush in the image space.
	\item Epidemiological socio-demographic or medical attribute data.
	%
	These values are associated with every subject represented in the image space. 
	%
	They do not describe shape information.
	%
	This is the data epidemiologists usually want to correlate with image data.
\end{itemize}
%Other data associated to elements in the object space are considered to be the \emph{Attribute Space Domain} and are displayed using various range based views using statistical graphics and information visualization techniques like scatterplots, bar charts or parallel coordinates \cite{Oeltze2013, Oeltze2007}.
%
%Specific for the epidemiological application domain is that both Object- and Feature Space Domain influence the outcome of interest.
% \textbf{Object Space}\\
% - Medical Image Data / Spine Segmentations\\
% \textbf{Attribute Space}\\
% - SHIP-Variables\\
% - Derived visualizations
% 
% ? Where to incorporate Levels of IVA

\subsection{IVA Patterns}
%
\emph{IVA} knows three different exploration patterns, handling associations of domain and range variables.
%The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handlinassociationon between domain and range variables.

\paragraph{Local Investigation}
This pattern projects information from domain space to the range perspective.
%
This step is more complicated in epidemiological context compared to other \emph{IVA} application domains.
%As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
%
Shape information can not be brushed using ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
%
\emph{Feature selection} methods strongly depend on the type of segmentation used to extract the tissue of interest.
%
Model-based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
%
Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013VMV}.
%
These algorithms can be used to investigate associations between shape-groups and other non-image based variables.
%
Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
%
%Another application is the outlier analysis. 
%
%Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.

% Local Investigation
% - Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
% - Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
% - Create a Pipeline Overview over different Levels of different IVA Patterns and Stages

% \subsection{Feature Selection}

% \subsection{Brushing the information space}
% - derive new groups which serve as input for the feature selection

\paragraph{Feature Localization}
% Subset selection of range variables yields features
The vast majority of features are considered to be dependent with respect to the image domain in the \emph{IVA} context.
%As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the \emph{IVA} context.
%
Selecting subjects based on image derived data can be seen as additional possibility of shape-related grouping.
%
The epidemiologist is interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
%
Epidemiologists may categorize data into groups that fit their hypothesis formulation.
%
Continuous variables such as age are for example often divided in categories like young, aged and elderly.
%
%Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in Section~\ref{sec:AdaptiveFeatureVisualization}.
%
%To gain insight intassociationons between SHIP-

% - Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
% - this selection is for categorical data already given implicitly bei projecting the 3D-View onto the Bar Charts/Mosaik Plots!\\
% - this aims to locale features of the data!\\
% - Gain Information about SHIP-Variables by putting them into the context of each other. The Pivot Table allows for direct numerical\\ analysis, while the information visualizations allow for better insight of the combination

\paragraph{Multivariate Analysis}
%Introduced in the information visualization, the multivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
%Introduced in the information visualization, the multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
Multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
%
The type of the parameters which are compared determines the information visualization technique used.
%
%This represents for the most part the usual approach which essentially is the analysis of how variables interact with each other, only in a visual analysis context.
%
Statistic measures are needed which describe how variables correlate given the selected groups.
%The need of statistic measures which describe how variables correlate with each other given the selected groups is special for the application domain.
%
These associations are also summarized using pivot tables which are popular in epidemiology.
% - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
% - Becker, R.A., Cleveland, W.S.: Brushing scatterplots. Technometrics 29(2) (1987)\\
% - Wang Baldonado, M.Q., Woodruff, A., Kuchinsky, A.: Guidelines for using multiple views in information visualization. In: AVI ’00: Proceedings of the working conference on Ad- vanced visual interfaces, pp. 110–119. ACM Press, New York, NY, USA (2000). DOI\\ http://doi.acm.org/10.1145/345513.345271
% - By brushing individual parameters or create new binnings of parameter it is possible to see how they change in coordinated views. This is already implemented in the Cargo framework\\
% - by creating comparative 3d Visualizations it is possible to assess the influence of non-image parameters to the visual space.

\subsection{Data Preprocessing} \label{Data Preprocessing}
Transformation operations on the data to prepare it for a \emph{IVA} system are denoted as data preprocessing.
\paragraph{Non-Image data. }
%To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
%
Multimodal features require different techniques.
%Due to the different acquisition modalities there have to be different techniques incorporated.
%
Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS} or \texttt{Stata}, which have a proprietary data format with limited export capabilities.
%
Exporting the data in the respective tool to a CSV file and then convert it to file types that are easily manageable such as JSON or XML makes it readable for all modern programming languages.
%The best solution for us was to simply export the data in the respective tool to a CSV file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
%
This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org/}}, which also validates the data (find missing data, clean up bad formatting, transform scales).
%In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
%
Including the data dictionary, which stores information about each manifestation of a feature, helps to get a detailed description of data variables and the meaning and unit of measurement of their values.
%Exporting the data dictionary, which stores information about each manifestation of a feature, is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
%
Missing data is denoted using Error codes indicating its cause ranging from ethical to medical and personal issues.
%The reasons for missing data have a wide range from ethical to medical and personal issues.
%
%Therefore, these are also included as error codes which have to marked as such in the data dictionary.
\paragraph{Image data. }
Processing the image data associated to each subject consists for the most part of information extraction about a anatomical structure.
%
This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as graph cuts \cite{GraphCut}) or by a (semi)-automatic detection, registration and segmentation.
%
These algorithms have to deal with a large inter-subject variability of the anatomical structure and need to produce reproducible results \cite{Preim2014}.
%
Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
%
If a segmentation yields only binary masks separating the structures, algorithms such as Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
%
Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
%
Grey value comparison is usually used to measure the quantity of fat, water, and--application specific--iron content (liver) or the distribution of grey and white brain tissue.

Morphometric variables are derived to allow for statistical comparison of the tissue which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed Version of the \emph{IVA Loop}. Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types. The data is visualized in the range and domain space, which can then be brushed, yielding new groups which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering which creates shape groups.
%
Hypothesis-generation with focus on image-data starts with a shape based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type.
%
The visualization techniques have to combine both image-and non image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new variables can be added to the analysis.
%
Brushes act as feature as they divide the subject space just like categorical variables and lead therefore to a new variable selection.
%
Selecting features also triggers a contingency analysis highlighting variables interacting with selected features, which is described in the following section.

%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
%
%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
%
A sample workflow using interaction- and visualization techniques described in the following section can be seed in Figure~\ref{fig:visualization}.

@@

\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Adjacency matrix of 131 features with a grand total of 17.161 combinations. Similarity is calculated using the \emph{Cram\'{e}rs' V} contingency value. Color brightness encodes association strength. The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 }
 \label{fig:similarity}
\end{figure}

>>>>

\documentclass[journal]{style/vgtc} 			          % final (journal style)
%\documentclass[review,journal]{style/vgtc}         % review (journal style)
%\documentclass[widereview]{style/vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{style/vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal
% TODO: Reference Nicolas Kruchten
%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data is analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{associations}).
%
Modern cohort studies also incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It shows to be a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.

@@

% \abstract{Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease specific risk factors.
% %
% Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
% %
% %Using well established statistical methods the data is hypothesis driven analyzed to find statistically significant variable correlations (\emph{associations}).
% Driven by a priori defined hypotheses, the data is analyzed using well established statistical methods in order to find statistically significant correlations between variables (\emph{associations}).
% %
% Modern cohort studies also incorporate medical image data.
% %
% Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
% %
% The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
% \\\\
% We propose an Interactive Visual Analysis approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
% %
% Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.
% %As extension to the epidemiological workflow it aims to unveil associations based on all reliable shape information rather than abstracting it to a few key figures.
% %Iterative analysis between ... allows for extracting associations which without the visiual analysis driven approach maybe would be overlooked. %Einsicht in Interaktionen zwischen nicht-bilddaten und bilddaten auf basis aller verfügbarer Informationen 
% %Iterative Analyse von Organstrukturen im Bezug auf Bilddaten 
% %Iterative analysis of image-based epidemiological datasets aims to provide insight into associations between organ shape and % in order to provide insights into \emph{associations} by incorporating additional information
% %It extends the epidemiological workflow with an iterative Analysis process and 
% It allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
% %
% Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
% %
% Similarity measures between data variables are used to compute interesting changes in variable associations for the current variable selection.
% %
% Shape based grouping of subjects is facilitated using clustering techniques which operate on surface meshes extracted from the image segmentation.
% } % end of abstract

@@

\paragraph{Domain and Range Variables. }
In the \emph{IVA} context, data is characterized by a combination of independent variables, such as space and/or time, and dependent variables, like temperature or pressure.
%
Two kinds of views are employed to inspect the data:
\begin{itemize} 
	\item \emph{physical views} \cite{Oeltze2013}, such as direct volume rendering, show information in the context of the spatiotemporal observation space \cite{Oeltze2007}, while
	\item \emph{attribute views}, such as scatter plots and parallel coordinates, show relationships between multiple data attributes.
\end{itemize}
Transferred to epidemiological data, the residential area of cohort subjects could be interpreted as space, the different assessment cycles of a longitudinal study as time, and the image and non-image data as dependent variables.
%
However, our current work neglects geographical and temporal aspects.
%
Instead we employ a more abstract model and consider the subjects as living in a joint image space where each of them is represented by a particular segmented organ or structure.
%
For instance, the spine is segmented over all subjects and all spines are co-registered spanning a joint space.
%
Two types of dependent variables then exist: 
\begin{enumerate}
	\item the socio-demographic data and medical examination results and 
	\item variables derived from the segmented structures, e.g., spinal curvature or misalignment of the vertebrae.
\end{enumerate}
An alternative of the image space would be the shape space generated by extracting the major modes of variation from all segmentation results \cite{Busking2010a}.
%
Based on our abstract model, the three analysis patterns of \emph{IVA} can be employed.

\paragraph{Local Investigation:} refers to the inspection of dependent variables with respect to certain subsets of the image or shape space.
%
For instance, the epidemiologist selects several spines with a common characteristic in the image or shape space and wishes to inspect the associated dependent variables in an attribute view \cite{Hermann2014}.
%
The selection step requires dedicated interaction techniques for defining a subset.
%
Alternatively, derived shape-related variables opposed in an attribute view or automatic techniques for shape clustering may be employed \cite{Klemm2013VMV}.
%
Clustering algorithms can be used to investigate associations between shape-groups and other non-image based variables.
%
Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.

\paragraph{Feature Localization:} refers to the search for structures in the image or shape space with a defined characteristic.
%
The epidemiologist may be interested in all female subjects with lower back pain and wishes to see the corresponding spines in a physical 3D view.

\paragraph{Multivariate Analysis:} refers to an investigation of multi-variate properties of the dependent data by specifying a feature in one attribute view and at the same time analyzing the values distribution with respect to other variables in other attribute views.
%
The epidemiologist may define a feature in a scatter plot of BMI and age and inspect the result in a histogram of body height.
%
These associations may also summarized using pivot tables which are popular in epidemiology.

% ---------------------------- old part ----------------------------
% In the \emph{IVA} context, data is divided into two major view types.
% The human body exposing shape information for the \emph{physical view} \cite{Oeltze2013}.
% This information space is usually displayed via volume rendering techniques \cite{Oeltze2007}.
% %Spatiotemporal views include data from the \emph{Object Space Domain} renders data centered to the focused domain, in our example the medical image data space.
% These variables spanning the 3D-space are also referred to as \emph{domain variables}.
% \emph{Range variables} in the epidemiological context can be divided twofold:
% \begin{itemize}
% 	\item Variables derived from the image data. 
% 	These measures quantify shape information to allow for comparison and can also be used to brush in the image space.
% 	%These variables describe image data and can also be used to brush in the image space.
% 	\item Epidemiological socio-demographic or medical attribute data.
% 	These values are associated with every subject represented in the image space. 
% 	They do not describe shape information.
% 	This is the data epidemiologists usually want to correlate with image data.
% \end{itemize}
% %Other data associated to elements in the object space are considered to be the \emph{Attribute Space Domain} and are displayed using various range based views using statistical graphics and information visualization techniques like scatterplots, bar charts or parallel coordinates \cite{Oeltze2013, Oeltze2007}.
% %Specific for the epidemiological application domain is that both Object- and Feature Space Domain influence the outcome of interest.
% % \textbf{Object Space}\\
% % - Medical Image Data / Spine Segmentations\\
% % \textbf{Attribute Space}\\
% % - SHIP-Variables\\
% % - Derived visualizations
% % ? Where to incorporate Levels of IVA
% \subsection{IVA Patterns}
% \emph{IVA} knows three different exploration patterns, handling associations of domain and range variables.
% %The explorative procedures when analyzing data using \emph{IVA} can be divided into three different patterns, handlinassociationon between domain and range variables.
% \paragraph{Local Investigation}
% This pattern projects information from domain space to the range perspective.
% This step is more complicated in epidemiological context compared to other \emph{IVA} application domains.
% %As opposing to other \emph{IVA} application domains, this step is more complicated in the epidemiological context.
% Shape information can not be brushed using ROI-selections but rather has to employ techniques that specify local deformation changes \cite{Hermann2014} or subjects that belong to a shape class.
% \emph{Feature selection} methods strongly depend on the type of segmentation used to extract the tissue of interest.
% Model-based segmentations or masks yield data structures capable of calculating mean shapes and distances between individuals or subject groups.
% Feature selection is also possible by applying clustering algorithms in order to get shape-groups \cite{Klemm2013VMV}.
% These algorithms can be used to investigate associations between shape-groups and other non-image based variables.
% Analysis of outliers can indicate segmentation errors or an group of subjects sharing a pathology.
% %Another application is the outlier analysis. 
% %Outliers can indicate segmentation errors or an outstanding group of individuals who may share a pathology.
% % Local Investigation
% % - Projection from Image Space to Attribute data. This is more difficult in our application, because we currently only brush on derived features.\\
% % - Clustering in Image Space yields Groups which can be analyzed using the attribute space\\
% % - Create a Pipeline Overview over different Levels of different IVA Patterns and Stages
% % \subsection{Feature Selection}
% % \subsection{Brushing the information space}
% % - derive new groups which serve as input for the feature selection
% \paragraph{Feature Localization}
% % Subset selection of range variables yields features
% The vast majority of features are considered to be dependent with respect to the image domain in the \emph{IVA} context.
% %As described before, the vast majority of data points are considered to be dependent with respect to the image domain in the \emph{IVA} context.
% Selecting subjects based on image derived data can be seen as additional possibility of shape-related grouping.
% The epidemiologist is interested in the shape of subjects within a range of a set of variables that describe the current hypothesis.
% Epidemiologists may categorize data into groups that fit their hypothesis formulation.
% Continuous variables such as age are for example often divided in categories like young, aged and elderly.
% %Categorization is strongly dependent on the hypothesis and therefore requires suitable brushing techniques as described in Section~\ref{sec:AdaptiveFeatureVisualization}.
% %To gain insight intassociationons between SHIP-
% % - Selection of Information using Bar Charts, Scatterplots or Parallel Coordinates which projects the selection into the Object space\\
% % - this selection is for categorical data already given implicitly bei projecting the 3D-View onto the Bar Charts/Mosaik Plots!\\
% % - this aims to locale features of the data!\\
% % - Gain Information about SHIP-Variables by putting them into the context of each other. The Pivot Table allows for direct numerical\\ analysis, while the information visualizations allow for better insight of the combination
% \paragraph{Multivariate Analysis}
% %Introduced in the information visualization, the multivariate analysis means selecting subjects in like in the feature localization step and get have the result highlighted in another linked view displaying other non-image parameter.
% %Introduced in the information visualization, the multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
% Multivariate analysis incorporates brushing and linking of of views displaying non-image parameter.
% The type of the parameters which are compared determines the information visualization technique used.
% %This represents for the most part the usual approach which essentially is the analysis of how variables interact with each other, only in a visual analysis context.
% Statistic measures are needed which describe how variables correlate given the selected groups.
% %The need of statistic measures which describe how variables correlate with each other given the selected groups is special for the application domain.
% These associations are also summarized using pivot tables which are popular in epidemiology.
% % - Selection of Elements in Scatterplots, Barcharts or Parallel Coordinates Views - observe how selection changes another view - this allows for multivariate analysis\\
% % - Becker, R.A., Cleveland, W.S.: Brushing scatterplots. Technometrics 29(2) (1987)\\
% % - Wang Baldonado, M.Q., Woodruff, A., Kuchinsky, A.: Guidelines for using multiple views in information visualization. In: AVI ’00: Proceedings of the working conference on Ad- vanced visual interfaces, pp. 110–119. ACM Press, New York, NY, USA (2000). DOI\\ http://doi.acm.org/10.1145/345513.345271
% % - By brushing individual parameters or create new binnings of parameter it is possible to see how they change in coordinated views. This is already implemented in the Cargo framework\\
% % - by creating comparative 3d Visualizations it is possible to assess the influence of non-image parameters to the visual space.
% ---------------------------- old part ----------------------------

\subsection{Data Preprocessing} \label{Data Preprocessing}
Transformation operations on the data to prepare it for a \emph{IVA} system are denoted as data preprocessing.
\paragraph{Non-Image data. }
%To include heterogenous epidemiological data in an \emph{IVA}-framework it is necessary to process it to obtain standardized views to the available features.
%
Multimodal features require different techniques.
%Due to the different acquisition modalities there have to be different techniques incorporated.
%
Data obtained using questionnaires or medical tests are often stored using statistical packages such as \texttt{SPSS} or \texttt{Stata}, which have a proprietary data format with limited export capabilities.
%
Exporting the data in the respective tool to a CSV file and then convert it to file types that are easily manageable such as JSON or XML makes it readable for all modern programming languages.
%The best solution for us was to simply export the data in the respective tool to a CSV file and then convert it to data types which are easier manageable such as JSON or XML using our own classes.
%
This can be achieved using data wrangling tools such as \texttt{OpenRefine}\footnote{Developed by Google, Open Source; \url{http://openrefine.org/}}, which also validates the data (find missing data, clean up bad formatting, transform scales).
%In order to verify that the conversion worked as expected and the data is valid, it is good practice to use data wrangling tools such as \texttt{OpenRefine} to validate the data (find missing data, clean up bad formatting).
%
%Including the data dictionary, which stores information about each manifestation of a feature, helps to get a detailed description of data variables and the meaning and unit of measurement of their values.
A data dictionary stores information about each manifestation of a feature and is stored as lookup table to get a detailed description of data variables and the meaning and unit of measurement of their values.
%Exporting the data dictionary, which stores information about each manifestation of a feature, is also an important step to get a detailed description of data variables and the meaning and unit of measurement of their values.
%
Missing data is denoted using Error codes indicating its cause ranging from ethical to medical and personal issues.
%The reasons for missing data have a wide range from ethical to medical and personal issues.
%
%Therefore, these are also included as error codes which have to marked as such in the data dictionary.
\paragraph{Image data. }
Processing the image data associated to each subject consists for the most part of information extraction about a anatomical structure.
%
This is either done manually by experts setting landmarks (sometimes supported by algorithms connecting the land marks such as graph cuts \cite{GraphCut}) or by a (semi)-automatic detection, registration and segmentation.
%
These algorithms have to deal with a large inter-subject variability of the anatomical structure and need to produce reproducible results \cite{Preim2014}.
%
Model-based approaches have shown in principle to be effective for segmentation \cite{Gloger2010, Gloger2012} and detection \cite{Rak2013}.
%
If a segmentation yields only binary masks separating the structures, algorithms such as Growing and Adaptive Shapes can be applied creating a surface grid where each point is comparable throughout the population \cite{Ferrarini2007}.
%
Intensity-based comparisons can be achieved using rigid image registration, but model based results however are preferable \cite{Klemm2012}.
%
Grey value comparison is usually used to measure the quantity of fat, water, and--application specific--iron content (liver) or the distribution of grey and white brain tissue.

Morphometric variables are derived to allow for statistical comparison of the tissue which incorporate mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 % TODO: Recreate Image: Also include Cramér´s V which is part of the visualization
 \caption{Detailed Version of the \emph{IVA Loop}. Usually starting with a selection of Feature of interest (user-driven or via data mining techniques), the data is mapped using a visualization technique appropriate for the selected data types. The data is visualized in the range and domain space, which can then be brushed, yielding new groups which can be investigated using further features. Note that adjacent steps are directly connected via feedback loops allowing for iterative refinement and give as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, -visualization and -brushing.
%
An hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection can also be derived from a shape-based clustering which creates shape groups.
%
Hypothesis-generation with focus on image-data starts with a shape-based clustering.
%
The feature is mapped using a automatically chosen visualization appropriate for its data type.
%
The visualization techniques have to combine both image-and non image data in order to set domain and range data in relation to each other.
%
Subsequently the visualization can be either brushed or new variables can be added to the analysis.
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushes act as feature as they divide the subject space just like categorical variables and lead therefore to a new variable selection.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight variables interacting with selected features.
%
%The starting point of \emph{IVA Loop} is determined whether the analysis is hypothesis-driven or exploratory.
%
%An analysis with the aim of hypothesis generation of image-based data starts with a clustering of the structure 
%
A sample workflow using interaction- and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

\begin{figure}[htb]
 \centering
 \includegraphics[width=3.5in]{figures/similarity_matrix}
 \caption{Adjacency matrix of 131 features with a grand total of 17.161 combinations.
 %
 Similarity is calculated using the \emph{Cram\'{e}rs' V} contingency value.
 %
 Color brightness encodes association strength.
 %Mouse-over focuses a entry and highlights the feature names for better readability.
 Mouser-over an entry highlight the feature names for better readability.
 %
 The enlarged excerpt shows associations for shape clusters of subjects with and without diagnosed spine attrition, which show associations between gender, weight, body height and smoking behavior.
 }
 \label{fig:similarity}
\end{figure}

\subsubsection{System Structure} \label{Structure and Workflow}
We divide the workspace into four major parts as seen in Figure~\ref{fig:visualization} and Figure~\ref{fig:similarity}.
\begin{itemize}
	\item The sidebar, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
	\item The canvas which holds all visualizations. Elements can be added, arranged, resized and removed freely.
	\item The interactive Pivot Table gives detailed numerical information of the features in the canvas view. This view on the data come natural to epidemiologists.
	\item The contingency view depicting relations for features in the canvas in an adjacency matrix.
\end{itemize}
%
% TODO Include:
% The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
% 
% Features can be dragged from the sidebar into the canvas area.
% %
% Doing so will automatically create an information visualization suitable for the current data type.
% %
% 3D shape information about the investigated structure is displayed for each variable manifestation.
% %
% By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
% %
% Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
% %
% Shape based clustering can either be applied to all subjects or subgroups.
% 
% All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
% %
% Details to the different views are presented in the following sub-sections.
